---
title: 'Prosper Loan Data Analysis by Jenny Hung'
output:
  html_document: default
  html_notebook: default
  pdf_document: default
Author: Jenny Hung
---

# Overview
In this project, we analyze the Propser Loan dataset. Our goal is to explore the data and discover patterns and relationships between selected variables - from an investor or lender's perspective.

Our goal for the project is to answer the following questions:

1. What are the factors that influences the BorrowersRate (conjectured: FICO, Rating, Term, Job v. HomeOwnership)
2. What are the factors that influences the LendersYield (conjectured: FICNAD, Rating, Term)
3. What are the relationship between Loan status and Lender's Yield (are they normally distributed, boxplot)
4. What are the relationship between # of investors, lendersYield, Rating/FICO, and number of recommendations
5. What about impact of job or ownership's on loan status, borrowers rate, and status

Before we continue, we first load the data and explore the structure of the data set and nature of variables. We see that the data set has 81 variables and contains 113,937 data points.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(GGally)
library(dplyr)
library(reshape2)
library(gridExtra)
```


```{r}
## first subset the data for ease of handling and exploration
loan = read.csv('prosperLoanData.csv', header=TRUE)
dim(loan)
```

The variables of the data set is listed below:

```{r}
colnames(loan)
```


# Preprocessing

```{r}
## Make ProsperRatingNumeric a factor variable
loan$ProsperRatingNumeric = factor(loan$ProsperRatingNumeric, levels=c(0, 1, 2, 3, 4, 5, 6, 7))
loan$ProsperRatingAlpha = factor(loan$ProsperRatingAlpha, levels=c('N/A', 'HR', 'E', 'D', 'C', 'B', 'A', 'AA'))
loan$Term = factor(loan$Term, levels=c(12, 36, 60))
```


```{r}
index = sample(nrow(loan), 1000)
```



```{r}
cols = c('CreditGrade', 'Term', 'LoanStatus', 'BorrowerRate', 'LenderYield', 'EstimatedLoss', 'EstimatedReturn', 
         'ProsperRatingNumeric', 'ProsperRatingAlpha', 'ProsperScore', 'ListingCategory..numeric.',
         'EmploymentStatus', 'EmploymentStatusDuration', 'IsBorrowerHomeowner', 'ProsperPrincipalBorrowed',
         'LoanOriginalAmount', 'Recommendations', 'Investors', 'DebtToIncomeRatio', 'StatedMonthlyIncome', 'MonthlyLoanPayment')

loan_subset = loan[index, cols]
```



# Univariate Plot Section

```{r}
numeric = ggplot(loan, aes(x=ProsperRatingAlpha)) + geom_bar()
numeric
```

We see that ProsperRatingAlpha is approximately normally distributed. Note that there are quite a few data points where the Prosper Rating is not available (as observed in NA values). 

```{r}
numeric = ggplot(loan, aes(x=EmploymentStatus)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
numeric
```

Most of the loans has an Employed as a job status.

```{r, message=FALSE, warning=FALSE, }
numeric = ggplot(loan, aes(x=EmploymentStatusDuration)) + geom_bar()
numeric
```

Even though a vast majority of loans has Employed as Employment Status, the employement duration in number of days is not normally distributed. In fact, it is skewed with a long right tail.

```{r}
numeric = ggplot(loan, aes(x=IsBorrowerHomeowner)) + geom_bar()
numeric
```

The home ownership status for successful loan appicants is approximately equally distributed.


```{r}
numeric = ggplot(loan, aes(x=Recommendations)) + geom_bar()
numeric
```

When looking at how many recommendations each loan received, we see that most of the loans received no recommendations. Note that this is the number of recommendations a borrower received at the time when the loan was created or listed.


```{r}
numeric = ggplot(loan, aes(x=Term)) + geom_bar()
numeric
```


```{r}
numeric = ggplot(loan, aes(x=LoanStatus)) + geom_bar() + theme(axis.text.x = element_text(angle = 90, hjust = 1))
numeric
```


```{r}
#TODO: BIN
numeric = ggplot(loan, aes(x=ProsperRatingAlpha)) + geom_bar()
numeric
```



```{r}
# TODO: BIN
numeric = ggplot(loan, aes(x=ProsperRatingAlpha)) + geom_bar()
numeric
```


```{r}
mean(loan$BorrowerRate)
```


```{r}
mean(loan$LenderYield)
```


```{r}
## Explore distribution of BorrowerState
state = ggplot(loan, aes(x=BorrowerState)) + geom_bar() + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1))
state


```



# Univariate Analysis

## Structure of data set

There are 113,937 loans in the data set and 81 features. Of the 81 features, The ProsperRating, which comes in two flavors, ProsperRatingAlpha and ProsperRatingNumeric, are both Prosper's own credit reporting scale, and it has the following structure:

(Either no credit or worst) ------> (best)
ProsperatingNumeric: 0, 1, 2, 3, 4, 5, 6, 7
ProsperatingAlpha: N/A, HR, E, D, C, B, A, AA
BorrowerRateThe Borrower's interest rate for this loan. 
LenderYield: The Lender yield on the loan. Lender yield is equal to the interest rate on the loan less the servicing fee.

Other observations:

Most loas are of 36-month term.
Most laon applicant are employed.
Approximately half of loan applicants are also home owners.
Most of the loan applicants have received no recommendations.
Mean BorrowRate is 19.28%.
Mean LenderYield is 18.27%.


## Main feature of interest

Due to the large number of variables, I have chosen to focus on BorrowerRate, LendersYield, Term, ProsperRatingNumeric/ProsperRatingAlpha, Recommendations, Investors, DebtToIncomeRatio, StatedMonthlyIncome, and MonthlyLoanPayment.

## Any created or constructed features

`BorrowerRate` and `LenderYield` are bucketed. This allowed me to see the distributions of these two variables more easily.

In addition, `PaymentIncomeRatio` is also constructed which is a ratio of `MonthlyLoanPayment` over `StatedMonthlyIncome`.

Careful exmination revealed that the missing values contained in the data set can be disregarded as they are most likely to be missing completely at random for the varialbes that are considered in the preceeding paragraph. This is because the nature of data set (a set of succesfuly approved loans) and their presence does not depend on observed or unobserved measurements of other features considered.


# Bivariate Plot Section

```{r}
## Explore correlation between CreidtGrade and and ProsperRatingAr
## Explore correlation between BorrowerRate and LendersYield
s1 = ggplot(data=loan, aes(x=BorrowerRate, y=LenderYield)) + geom_point(alpha = 0.3)
s1
```

Not surprisingly, the scatter plot of BorrowerRate and LenderYield approximately fall on a stright line passing through the origin. The following calcualtion also show that these two variables are of perfect correlation.

```{r}
with(loan, cor.test(BorrowerRate, LenderYield))
```


```{r}

### Major finding
rate_groups = group_by(loan, ProsperRatingAlpha)
rate_rec = summarise(rate_groups, 
                      recommendation_mean = mean(Recommendations),
                      recommendation_median = median(Recommendations),
                      recommendation_sd = sd(Recommendations))
rate_rec = arrange(rate_rec, ProsperRatingAlpha)

melted = melt(rate_rec, id='ProsperRatingAlpha')
```

```{r}
insight1 = ggplot(melted, aes(ProsperRatingAlpha, y=value, color=variable, group=variable)) + geom_line() +
    geom_point() + ylab('Recommendations of Loan')
insight1
```

The above graph shows the relationship between the credit rating of the borrowed against the number of recommendations that a borrower received for the loan. The mean, median as well as standard deviation of the number of recommendations are shown. Not surprising, the median recommendation is most robust, as it barely moved at all, whereas the mean had shifted based on the credit raitng of the borrower.

```{r}
## A summary table by # of Recommendation, mean yield, and term
## Major finding 
rec_groups = group_by(loan, Recommendations)
rec_rate = summarise(rec_groups,
                     mean_yield = mean(LenderYield),
                     median_yield = median(LenderYield),
                     term12 = sum(Term == 12),
                     term36 = sum(Term == 36),
                     term60 = sum(Term == 60))
rec_rate = arrange(rec_rate, Recommendations)

```

```{r}
insight2 = ggplot(rec_rate, aes(factor(Recommendations), y=mean_yield, group=1)) + geom_point() + geom_line() +
    xlab('Number of Recommendations') + ylab('Mean LenderYield')
insight2
```


The above graph shows the relationship between the numeber of recommendations vs. the mean LendersYield. 


```{r, message=FALSE, warning=FALSE}
## Explore relationship between PropserRatingAlpha, LenderYield, and Term
## Major finding
rel_basic = ggplot(data=loan_subset, aes(x=ProsperRatingAlpha, y=LenderYield)) + geom_boxplot() 
rel_basic
```




# Bivariate Analysis

## Major relationships observed

* There is a near perfect correlation between BorrowerRate and LenderYield. 
* In general, BorrowerRate is inversely proportional to the credit rating of a borrower. Given the near perfect correlation, this also means that LenderYield is inversely proportional to the credit raitng of a borrower.
* In cases where a borrower has no established credit rating (namely, where `ProsperRatingNumeric` or `ProsperRatingAlpha` is NA), the LenderYield tend to be somewhere in between the interest rates for the loans with the best credit and with the worst credit.


## Surprising finding
* Previously, I had conjectured that the LendersYield is correlated to the number of recommendations that a borrower receives. However, this prior belief does not bear out in the exploratory analysis. In fact, the type of borrowers who receive the highest number of recommendations is one that does not have a credit rating, followed immediately by borrowers who have the best credit rating. 


## The strongest relationship
* The strongest relationship found is the invserse relationship between the interest rate at which a borrower pays and the borrower's credit strength.



# Multivariate Plot Section

```{r, message=FALSE, warning=FALSE}
## Explore relationship between PropserRatingAlpha, LenderYield, and Term
## Major finding
rel = ggplot(data=loan_subset, aes(x=ProsperRatingAlpha, y=LenderYield)) + geom_boxplot() +
    facet_wrap(~Term)
rel
```

The faceted boxplot shows the distribution of credit rating and lender's yield across all loan terms. Here we see that the inverse relationship between ProsperRatingAlpha and LenderYield persists across all terms.


```{r}
## Investigate the relationship between LendersYield and DebtIncome ratio
## Find correlation r - low
with(loan, cor.test(DebtToIncomeRatio, LenderYield))
```

```{r}

## use this graph
debt_income_log = ggplot(subset(loan_subset, !is.na(DebtToIncomeRatio)), aes(DebtToIncomeRatio, LenderYield)) + 
    geom_point(alpha=0.05) + scale_x_log10() + scale_y_log10() +
    facet_wrap(~ProsperRatingNumeric)
debt_income_log
```

In the proceeding graph, we show the relationship between DebtToIncomeRatio and LenderYield across all PropserRatingNumeric (credit ratings). We have also performed feature transformation and show the graph on a log-log basis. To make the graph more easily readable, we have shown the graph on a subset of the loan data but the relationship is representative to the entire dataset due to the randomization of the seletion process.

Note that as the scatter at the bottom left corner represent the cluster of loans with the lowest `DebtToIncomeRatio` and the lowest `LenderYield`. As the scatter moved upwards, it indicates that `LenderYield` is also moving higher. Note that the cluster of points indicating no credit ratings have the highest level of variability. 


```{r}

## Investigate the relationship between LendersYield and Payment/Income ratio
## Find correlation r
loan['PaymentIncomeRatio'] = loan['MonthlyLoanPayment'] / loan['StatedMonthlyIncome']
loan['PaymentIncomeRatio'][loan['PaymentIncomeRatio']==Inf] = NA


with(subset(loan, !is.na(PaymentIncomeRatio)), cor.test(PaymentIncomeRatio, LenderYield))
```

```{r, message=FALSE, warning=FALSE}
pmt_income = ggplot(subset(loan, !is.na(PaymentIncomeRatio)), aes(PaymentIncomeRatio, LenderYield)) + 
    geom_point(alpha=0.05) + facet_wrap(~ProsperRatingNumeric) + scale_x_continuous(limit=c(0,1)) +
    scale_x_log10() + scale_y_log10()
    
pmt_income
```

In the proceeding graph, we show the relationship between PaymentIncomeRatio and LenderYield across all PropserRatingNumeric (credit ratings). We have also performed feature transformation and show the graph on a log-log basis. To make the graph more easily readable, we have shown the graph on a subset of the loan data but the relationship is representative to the entire dataset due to the randomization of the seletion process.

Note that as the scatter at the bottom left corner represent the cluster of loans with the lowest `PaymentIncomeRatio` and the lowest `LenderYield`. As the scatter moved upwards, it indicates that `LenderYield` is also moving higher. Note that the cluster of points indicating no credit ratings have the highest level of variability. 



```{r}
## Conclude whether DebtIncome ratio is highly correlated with PaymentIncoem ratio
## Find correlation r
with(loan, cor.test(DebtToIncomeRatio, PaymentIncomeRatio))
```

# Multivariate Analysis

* There is a near perfect correlation between BorrowerRate and LenderYield. 
* In general, BorrowerRate is inversely proportional to the credit rating of a borrower. Given the near perfect correlation, this also means that LenderYield is inversely proportional to the credit raitng of a borrower.
* The invsere relationship between BorrowerRate/LenderYield and credit rating is invariant in general across all terms.
* In addition, there is also an inverse relationship between BorrowerRate/LenderYield and DebtToIncomeRation (as well as PaymentIncomeRatio), and this pattern also holds across all levels of credit rating. However, the patterns is not clear in cases where a borrower has no established credit rating.


## The strongest relationship
* The strongest relationship found is the invserse relationship between the interest rate at which a borrower pays and the borrower's credit strength.




# Final Plots and Summary


```{r, message=FALSE, warning=FALSE}
## Explore relationship between PropserRatingAlpha, LenderYield, and Term
## Major finding
rel = ggplot(data=loan_subset, aes(x=ProsperRatingAlpha, y=LenderYield)) + geom_boxplot() +
    facet_wrap(~Term) + labs(title='Relationship by ProsperRatingAlpha and LenderYield by Terms')
rel
```



```{r}

## use this graph
debt_income_log = ggplot(subset(loan_subset, !is.na(DebtToIncomeRatio)), aes(DebtToIncomeRatio, LenderYield)) + 
    geom_point(alpha=0.05) + scale_x_log10() + scale_y_log10() +
    facet_wrap(~ProsperRatingNumeric) + 
    labs(title='Relationship by DebtToIncomeRatio and LenderYield by ProsperRatingsNumeric')
debt_income_log
```



```{r}
insight1 = ggplot(melted, aes(ProsperRatingAlpha, y=value, color=variable, group=variable)) + geom_line() +
    geom_point() + ylab('Recommendations of Loan') + labs(title='Credit Rating v. Number of Recommendations')
insight1
```

# Conclusion and Summary


